---
title: "Ch. 13 HW due 2.20.22"
author: "Brenda Onyango"
date: "2/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 13


### Errors and residuals 

Y = 2 + 3x + $\epsilon$

a. find error for observation A. 

9 = 2 + 3(2) + $\epsilon$

$\epsilon$ = 1

b. Y = $\beta_0$ + $\beta_1$*X + $\epsilon$

find the residual for Observation A 
9 = 1.9 + 3.1(2) + $\epsilon$

$\epsilon$ = 0.9 = residual since the residual is the difference between the prediction we make with our fitted line and the actual value. The "error term" represents the residual when we're using a prediction line. 

### Writing the regression equation 

2) I would use:

Y = B_0 + B_1*X + B_2*A + B_3*B + e

Including coefficients for A and B closes the backdoors X <- A -> Y and X <- B -> Y. C is included in e. 

### Coefficients 

3) 
a. The coefficient that multiplies a variable is a slope, so a one-unit increase in X results in a 1.3 increase in Y. 

b. 
```{r}
pnorm(q = (3/1.3)) #using pnorm and z-score to find the percentiles. Z score is score - mean/stdev and in this case we can use 0 as mean and the stadard error as stdev. 
```

The estimate of 3 is in the 98.95th percentile of the normal distribution. 

```{r}
a  <- (100 - 98.95)*2 
a
```

The above shows that finding something as far from 0 as the estimate 3 happens 2.09% of the time. I multiplied by 2 above to do a two-tailed test. If our alpha = 0.05 then we reject the null since 2.09% is less than 5% and the estimate is statistically significant from 0. 

c. To say that the coefficient is statistically significant from 0 is to say that 0 is an unlikely value for the coefficient. It doesn't mean that 3 is a true value or that that the effect is meaningful. 

### Women's work 

1) annual hours worked = years of education + constant
2) annual hours worked = #ofchildren + constant
3) annual hours worked = years of education +  #ofchildren + constant

are a basic representation of what the columns in the table represent. 

a. Adjusting for number of children a one-unit increase of years of education is associated with 76.185 increase in annual hours worked. 

b. When not adjusting for years of education, the standard error on the "children under 5" is 19.693. 

c. For a woman with 0 children and 0 years of education, the predicted number of hours worked annually is the constant or intercept which is 306.553. 

d. 3382 observations are used in each regression. 

e. The coefficient on "children under 5" is statistically significant from 0 at the p<0.01 or the 99% level as indicated by the three asterisks. This also means that p is less than 0.05 so the coefficient is also statistically signicant from 0 at the 95% level. 