---
title: "Ch. 14 HW due 2.27.22"
author: "Brenda Onyango"
date: "2/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 14 Homework: Matching

### Matching with Penmanship Data

1) a. Create a set of weights for handedness groups so that treatment matches control.

Like in 4.2 I will give a weight of 1 to everyone who is *treated.* The weights I give to the control group will fill in the spaces below. 

10*__ = 6 for left 

88*__ = 90 for right

2*__ = 4 for ambidextrous 

weights for control groups:

```{r}
leftweightcon <- 6/10
rightweightcon <- 90/88
ambweightcon <- 4/2

leftweightcon
rightweightcon
ambweightcon
```

b. 
```{r}
leftweighttreat <- 10/6
rightweighttreat <- 88/90
ambweighttreat <- 2/4

leftweighttreat 
rightweighttreat 
ambweighttreat

```

c. Use the weights in part b to calculate proportion of left-handed and other people in the treated group.

```{r}
lefttreatprop <- 6*leftweighttreat
rightreatprop <- 90*rightweighttreat
ambidextprop <- 4*ambweighttreat #this was number of observations times weight to find proportion

lefttreatprop 
rightreatprop 
ambidextprop 
```

d. weighted average penmanship in treated group 
```{r}
# doing sum of weights times observation divided by sum of weights 

wapt <- (leftweighttreat*7 + rightweighttreat*6 + ambweighttreat*4)/(leftweighttreat + rightweighttreat + ambweighttreat)

wapt
```

e. The unweighted treatment effect on the treated is:

```{r}
unweightedteffect <- (.06*7) + (.9 * 6) + (.04*4)
unweightedteffect
```

```{r}
unweightedatt <-unweightedteffect - 5 #5 is average penmanship of people who don't practice 
unweightedatt
```

If we use our weighted data then our effect of practicing cursive is the below ono a 10 point scale:

```{r}
practiceffect <- wapt - 5 #5 is the average penmanship of people who don't practice 
practiceffect
```

### 2: Types of Matching 

a. kernel matching 

b. k-nearest-neighbor matching with k = 2 

c. propensity score matching 

d. one-to-one distance matching

### 3: Bias v. Variance 


