---
title: "Class practice"
author: "Brenda Onyango"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
p <- seq(0,1, length.out = 50) #sequence of 0 to 1 of 50 numbers
lik <- dbinom(1,2,p)
plot(p, lik)
```

```{r}
pp <- seq(0,1, length.out = 50) #sequence of 0 to 1 of 50 numbers
lik <- dbinom(4,6,p) #applied to 6 and got 2
plot(p, lik) #this is plotting p 
```
## Using GSS data find Pr(college), Pr(college|one parent with degree), and Pr(college|<one parent) and Pr(college|x) = max

```{r}
library(tidyverse)
library(gssr)
```
```{r}
data(gss_all)
gss2018 <- gss_all |> 
  filter(year == 2018) #didn't have janitor package so just filtered by year 
  
```

## looking at data 
```{r}
data(gss_doc)
my_varnames <- colnames(gss2018)
gss_doc <- gss_doc |> filter(id %in% my_varnames) 
```

```{r}
d <- gss2018 |> select(degree) |> mutate(college = if_else(degree >= 3, 1, 0))
```

### probability of college
```{r}
sum(d$college)/nrow(d) #total probability
```

## probabilty of college given one parent has a degree 
```{r}
d <- gss2018 |> select(madeg, padeg) |> mutate(parents= case_when(madeg >= 3~ 2, padeg >= 3~ 4))
```
```{r}
glimpse(d)
```

```{r}
sum(d$parents,na.rm = T)/nrow(d) #T is indicating as long as its not na, not saying that if it's a number. know the difference
```

