---
title: "HW 10"
author: "Brenda Onyango"
date: "10/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 9.1

a) A normal prior is reasonable for B0 and B1 because these can take values around a center value (positive and negative or even zero).  

b) A normal prior isn't reasonable for $\sigma$ because sigma can't be negative. 

c) The difference between weakly informative priors and vague priors is that weakly informative priors are more focused in that they reflect uncertainty  
across a range of *sensible* parameter values, whereas vague priors put weight on non-sensible parameter values. With weakly informative priors, chains don't have to spend time exploring non-sensible parameter values. 

## Exercise 9.2

Predictor = X and response variable = Y

a) X = arm length, Y = height 
b) X = distance b/w home and work, Y = person's carbon footprint
c) X = age, Y = child's vocabulary level
d) X = person's sleep habits, Y = reaction time 

Notice these have an order i.e. a person's carbon footprint can't happen before they commute.

## Exercise 9.3 

Y = B0 + B1X where B0 is y-intercept (what typical Y value is when and X = 0) and B1 is slope. X indicates predictor.

a) Y_kangheight = B_0 + X_ageB_1 Here B_0 is the height of the kangaroo at 0 months old or when it's born and B_1 is how much the kangaroo grows for each increase in month. B_1 is positive.

b) Y_Githubfollowers = B_0 + X_commitsB_1 Here B_0 is the number of followers when the data scientists has 0 Github commits and B_1 is how many followers are added for each increase in Github commits. Assuming B_1 is positive. 

c) Y_visitors = B_0 + X_rainfallB_1 B_0 is the number of visitors to a local park when there are 0 inches of rainfall. B_1 is the decrease in visitors for each additional inches of rainfall. B_1 is negative.

d) Y_netflix = B_0 + X_sleepB_1 B_0 is the daily hours of Netflix a person watches when they have 0 hours of sleep. B_1 is the decrease in netflix hours for each increase in the hours of sleep. B_1 is negative. 

## Exercise 9.4 

$\sigma$ tells us about the spread of data in our sample. More spread or a higher standard deviation means that the relationship between X and Y is weaker. Less spread or lower variation means the relationship between X and Y is stronger. 

```{r}
# Load packages
library(bayesrules)
library(tidyverse)
library(rstan)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(janitor)
library(broom.mixed)
```

a) 