---
title: "HW 3"
author: "Brenda Onyango"
date: "9/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# HW 3: Probability Workshop

### Preparing for exercises. 
First I loaded tidyverse.
```{r}
library(tidyverse)
```
Next I create the data frame of a deck of cards. 
```{r}
deck <- tibble(
  card = rep(
    c("ace",
      "deuce",
      "three",
      "four",
      "five",
      "six",
      "seven",
      "eight",
      "nine",
      "ten",
      "jack",
      "queen",
      "king"
    ),4),
  suit =
    rep(c(
      "diamonds", 
      "hearts", 
      "clubs", 
      "spades"),
        each = 13),
  color = rep(c("red", 
                "black"),
              each = 26)) #dataframe of deck of cards
```
Next I viewed the deck of cards using glimpse.
```{r}
glimpse(deck) #deck has 52 cards and 3 variables.
```
## Question 1

What is the probability of getting one black card first and then one red card in two draws with replacement? 
First I found the probability of drawing a black card. 
```{r}
pr_black <- deck |> filter(color == "black") |> nrow()/nrow(deck)
```
Then I found the probability of drawing a red card. 
```{r}
pr_red <-  deck |> filter(color == "red") |> nrow()/nrow(deck)
```
Then I multiplied the probabilities of each individual event. 
```{r}
pr_black * pr_red
```
The probabily of getting a black card and then a red card with replacement is 0.25. 

### Question 2
For question 2, I first used ?sample_n in the console to understand the syntax for sample_n(). Then I repeated the chunk 5 times to generate 5 randomly sampled hands of 5 cards. 
```{r}
sample_n(deck, 5, replace = FALSE) #sample 5 rows of the deck without replacement
```
```{r}
sample_n(deck, 5, replace = FALSE) #2nd random hand 
```
```{r}
sample_n(deck, 5, replace = FALSE) #3rd random hand
```
```{r}
sample_n(deck, 5, replace = FALSE) #4th random hand
```
```{r}
sample_n(deck, 5, replace = FALSE) #fifth random hand
```
#### Question 2 part 2

What is the total number of *different* hands you could draw in poker? This question is asking from 52 cards, how many groups of 5 can I form. I will use a combination function to solve this. 
```{r}
choose(52, 5) # this function performed 52C5 = (52!)/(5!(52-5)!)
```
### Question 3
What is the probability that each suit is represented in one hand (5 cards) of poker?

First I checked the deck and saw each of the four suits has 13 cards.I assume we're drawing cards to form a hand of 5 without replacement. 

There are 4 wa



0