---
title: "HW 3"
author: "Brenda Onyango"
date: "9/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# HW 3: Probability Workshop

### Preparing for exercises. 
First I loaded tidyverse.
```{r}
library(tidyverse)
```
Next I create the data frame of a deck of cards. 
```{r}
deck <- tibble(
  card = rep(
    c("ace",
      "deuce",
      "three",
      "four",
      "five",
      "six",
      "seven",
      "eight",
      "nine",
      "ten",
      "jack",
      "queen",
      "king"
    ),4),
  suit =
    rep(c(
      "diamonds", 
      "hearts", 
      "clubs", 
      "spades"),
        each = 13),
  color = rep(c("red", 
                "black"),
              each = 26)) #dataframe of deck of cards
```
Next I viewed the deck of cards using glimpse.
```{r}
glimpse(deck) #deck has 52 cards and 3 variables.
```
## Question 1

What is the probability of getting one black card first and then one red card in two draws with replacement? 
First I found the probability of drawing a black card. 
```{r}
pr_black <- deck |> filter(color == "black") |> nrow()/nrow(deck)
```
Then I found the probability of drawing a red card. 
```{r}
pr_red <-  deck |> filter(color == "red") |> nrow()/nrow(deck)
```
Then I multiplied the probabilities of each individual event. 
```{r}
pr_black * pr_red
```
The probabily of getting a black card and then a red card with replacement is 0.25. 

### Question 2
For question 2, I first used ?sample_n in the console to understand the syntax for sample_n(). Then I repeated the chunk 5 times to generate 5 randomly sampled hands of 5 cards. 
```{r}
sample_n(deck, 5, replace = FALSE) #sample 5 rows of the deck without replacement
```
```{r}
sample_n(deck, 5, replace = FALSE) #2nd random hand 
```
```{r}
sample_n(deck, 5, replace = FALSE) #3rd random hand
```
```{r}
sample_n(deck, 5, replace = FALSE) #4th random hand
```
```{r}
sample_n(deck, 5, replace = FALSE) #fifth random hand
```
#### Question 2 part 2

What is the total number of *different* hands you could draw in poker? This question is asking from 52 cards, how many groups of 5 can I form. I will use a combination function to solve this. 
```{r}
a <- choose(52, 5) # this function performed 52C5 = (52!)/(5!(52-5)!)
choose(52,5)
```

### Question 3
What is the probability that each suit is represented in one hand (5 cards) of poker?

First I checked the deck and saw each of the four suits has 13 cards.I assume we're drawing cards to form a hand of 5 without replacement. One suit will be drawn twice. 

```{r}
s <- choose(4,1) #Choosing one suit out of 4 to be drawn twice
ss <- choose(13,2) #choosing 2 from one suit with 13 cards 
```
Multiplying "s" and "2s" by 13 x 13 x 13 gives the number of hands with 1 card from each suit. 
```{r}
s *ss * (13^3) #cubing 13 because 13 choose 1 is done 3 times for each of the other cards in the 5-card suit
```
To find the probability I will use the above as the numerator and the number of total possible hand outcomes as the denominator. 
```{r}
(s*ss*(13^3))/a
```

### Question 4

What is the probability of getting at least 1 ace in one 5-card hand? 
First I will find the probability of not getting aces. There are 4 aces in a 52-card deck. Without replacement the probability of drawing other cards 5 times is below. 
```{r}
not_ace <- (48/52)*(47/51)*(46/50)*(45/49)*(44/48)
```
Subtracting the probability of not_ace from 1 will give the probability of at least one ace. 
```{r}
1 - not_ace
```
Another way I could find this is by using a combination of 48 choose 5. 
```{r}
no_ace <- choose(48,5) #number of possible hands with no aces
```
Then using object "a" (the number of possible 5-card hands) as the denominator to find probability of no aces. 
```{r}
prob_noace <- (no_ace/a) #probability of hands with no aces 
```
Then subtract this from 1 to find the probability of a hand with at least 1 ace. 
```{r}
1 - prob_noace #complement rule to find probability of at least 1 ace
```

### Question 5 
To prepare for this question I enter the info from stackoverflow.
```{r}
df <- tibble(
  id = 1:6,
  name = c("Alice", "Bob", "Charlie", "Diane", "Fred", "Jenny")
) #creating dataframe with 6 names for the committee
```
Then I using crossing (de-duplicates and sorts inputs) and unite (pastes multiple columns into one) to create a dataframe of possible pairs for the committee. 
```{r}
df_pairs <- df |> 
  rename(id1 = id, name1 = name) |>  
  crossing(df |> 
             select(id2 = id, name2 = name)) |> 
  filter(id1 < id2) |> 
  unite(id, id1, id2) |> 
  unite(name, name1, name2)

df_pairs
```
I could also find the number of combinations possible using  choose().
```{r}
choose(6,2)
```

### Question 6
Modify the code above so it shows all combinations when order does matter.

```{r}
df_orderpairs <- df |> 
  rename(id1 = id, name1 = name) |>  
  crossing(df |> 
             select(id2 = id, name2 = name)) |> 
  filter(id1 != id2) |> #changed filter from earlier code to find all pairs in which names don't equal each other to find all ordered pairs
  unite(id, id1, id2) |> 
  unite(name, name1, name2)

df_orderpairs
```
Analytically I can perform a permutation.
```{r}
factorial(6)/factorial(4) #the denominator is correcting for counting pairs with same name e.g. Fred_Fred and equals (6 - 2)!
```

### Question 7 

A) 1000 draws of a fair die with a plot. 
First I will create a die object. 
```{r}
die <- 1:6
  
```
Next I will roll the die 1000 times  using the sample() function and set.seed() to save the rolls.  
```{r}
set.seed(1)
replicate(1000, sample(die,1)) |> hist() #roll 1 die and repeat 1000 times 

mean(replicate(1000, sample(die,1))) #gives mean of 3.48

```

B) A 1000 draws of a die which lands on 6 for 50% of the time. 

I will start by making a weighted die with probabilities for each value inclusive of 1 through 6.

```{r}
weighted_die <- 1:6
prob = c(".1", ".1",".1",".1", ".1", ".5")

```

Next I will roll weighted_die 1000 times and create a histogram. 

```{r}
set.seed(2)
replicate(1000, sample(weighted_die,1)) |> hist() #roll 1 weighted die and repeat 1000 times 

mean(replicate(1000, sample(weighted_die,1))) #gives mean of 3.441
```

C) For C, I will repeat the steps in B with different probabilities. 

```{r}
weighted_die2 <- 1:6
prob = c("(1/6)", "(1/6)","(1/6)","(1/6)", "(1/3)", "(1/3)")

set.seed(4)
replicate(1000,sample(weighted_die2,1)) |> hist()

mean(replicate(1000,sample(weighted_die2))) #calculating mean


```

## Citation 
Used the following link to think about question 3: https://math.stackexchange.com/questions/3670963/probability-calculating-the-probability-of-choosing-5-cards-with-4-different-s
