---
title: "HW 7_Bayes Ruels Chapter 4"
author: "Brenda Onyango"
date: "10/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### loading packages needed for the assignment. 

```{r}
library(tidyverse)
library(bayesrules)
```


## Exercise 4.1

a) Beta(1.8, 1.8) centers $\pi$ on 0.5
b) Beta(3,2) somewhat favors $\pi$ > 0.5 since alpha is slightly larger than beta
c) Beta(1,10) strongly favors  $\pi$ < 0.5
d) Beta(1,3) somewhat favors  $\pi$ < 0.5
e) Beta(17,2) strongly favors  $\pi$ > 0.5

## Exercise 4.2
We can eliminate a and b because the prior is not symmetrical or centered around .5. The prior is right-tailed meaning beta > alpha so we can eliminate f. This leaves c,d, and e which have the same Beta distribution but different y and n. 
The liklihood is symmetrical around pi of 0.50 so the correct choice is e where y is half of n. 
Below I've plotted e to show it matches. 
```{r}
plot_beta_binomial(alpha = 3, beta = 8, y = 2, n = 4)
```


## Exercise 4.3

a) Beta(3, 10) this gives a right-skewed distribution where pi is lower to indicate it's really unliklely. 
b) Beta(1,1) this gives a uniform distribution where all pi values between 0 and 1 are equally likley.
c)Beta(10,3) this gives a left-skewed distribution where pi is higher to indicate it's really likely. 
d)Beta(5,4) gives a distribution that slightly favors pi > 0.50 to indicate Daryl being slightly unsure. 
e)Beta(4,5) gives a distribution that slightly favors p < 0.50 to indicate Scott being slightly unsure. 

## Exercise 4.4

Kimya's prior indicates that she thinks there's a chance that the shop is closed but she is unsure. 
```{r}
plot_beta(alpha = 1, beta = 2)
```
Fernando's prior. Fernando thinks that it is more liklely the shop is closed, but is unsure. He is more sure it's closed than Kimya. 
```{r}
plot_beta(0.5, 1)
```
Ciara's prior. Ciara is very sure the shop is *not* open. 
```{r}
plot_beta(alpha = 3, beta = 10)
```

Taylor's prior. Taylor is sure the shop is open. 
```{r}
plot_beta(alpha = 2, beta = 0.10)
```

## Exercise 4.4 

We are given y = 3 and n = 7 and we will use these values and  pi|(Y=y)~Beta(alpha + y, B+ n- y) to simulate the posterior for each of the above models. We'll set y equal to z and n equal to q to avoid confusion. 
```{r}
z <- 3
q <- 7
#Kimya's model
plot_beta_binomial(alpha = 1, beta = 2, y = (1+z), n = (2 + q - z)) + geom_vline(xintercept = .5555556, color = "purple") 
summarize_beta_binomial(alpha = 1, beta = 2, y = (1+z), n = (2 + q - z))

#Fernando's model
plot_beta_binomial(alpha = 0.50, beta = 1, y = (.5+z), n = (1 + q - z)) + geom_vline(xintercept = 0.6153846, color = "purple")
summarize_beta_binomial(alpha = 0.50, beta = 1, y = (.5+z), n = (1 + q - z))

#Ciara's model
plot_beta_binomial(alpha = 3, beta = 10, y = (3+z), n = (10 + q - z)) + geom_vline(xintercept = 0.3333333, color = "purple")
summarize_beta_binomial(alpha = 3, beta = 10, y = (3+z), n = (10 + q - z))

#Taylor's model
plot_beta_binomial(alpha = 2, beta = 0.10, y = (2+z), n = (.10 + q - z)) + geom_vline(xintercept = 1.129032, color = "purple")
summarize_beta_binomial(alpha = 2, beta = 0.10, y = (2+z), n = (.10 + q - z))
```
```{r}
hist(plot_beta(

