---
title: "Soc. 723 HW due 1.16.22"
author: "Brenda Onyango"
date: "1/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 3 Homework

1) What is a variable? A variable is a characteristic that takes on different values across a unit of analysis. In the chapter monthly incomes of 433 South Africans was given as a variable, but it would be more precise to say that monthly income is the variable measured at the unit of the individual and that there are 433 observations for this variable. 

2) For each of the following variables describe if it is continuous, count, ordinal, categorical or qualitative. 

a) Age - continuous*
b) gender - categorical
c) number of times of presidential tweets - count 
d) income - continuous 
e) number of Insta posts about stats - count
g) college students' attend - categorical
h) therapist's written assessment - qualitative 
i) soccer divisions - ordinal 

*Technically continuous but I know practically for data analyses age is often treated as discrete. 

3) distribution

### Frequency Table and Salaries

Before answering 4, I created an object with given data to do some visualization and calculate relative frequencies. 

```{r}
#create table with salaries and frequency

salaryset <- c(rep('85K', 5), rep('90K', 4), rep('100K', 1), rep('120K', 2), 
rep('125K', 3), rep('130K', 2))

salarytable <- table(salaryset)

```

Let's look at the table.
```{r}
salarytable #notice that R arranged in different order than what was entered in set 
```

```{r}
#constructing relative frequency distribution of the data in the table.

s <- sum(salarytable)

#divide by sum to get relative frequencies
relfreq <- salarytable/s 

#view freq
relfreq
```

```{r}
dotchart(sort(relfreq),
         xlab = 'Relative frequency of salaries',
         main = 'Dot Chart of Econ Professor Salaries',
         xlim = c(0, 0.4), pch = 20, color = 'black')
```

a) Calculate average salary.
```{r}
salaries <- c(rep(85000, 5), rep(90000, 4), rep(100000, 1), rep(120000, 2), 
rep(125000, 3), rep(130000, 2))

head(salaries, 9) #confirming data entered correctly. Note that salaries is numerical and salaryset is categorical.
```

```{r}
mean(salaries)
```

Average salary is $103,529.40.

b) Find median.
```{r}
median(salaries)
```

The median is $90,000. 

c) Find min and max. 
```{r}
min(salaries)
max(salaries)
```
Min is 85,000 & max is $130,000. Instead of using min() and max(), I could have looked at the table. 

d) Find IQR. 
```{r}
IQR(salaries) #IQR is 40,000
summary(salaries) #quartiles 
```

5) c. Greek letters

6) a. Log transforming the variable 

7) Order from *low* to *high* variability: D, A, C, B  

8) 

a) The distribution ranges from about 73 about 89 with a mode of about 79. 

b) There is a slight right skew. 

c) Because of the right skew (very few people scoring very well), the median may be a better measure for the center of distribution.

d) I would use standard deviation to describe variability. 

9) b. The mean< median. 

### Frequency Table and Students 

```{r}
#create table with grade and frequency

students <- c(rep('First Year', 1000), rep('Sophomore', 1200), rep('Junior', 900), rep('Senior', 1500))

studenttable <- table(students)
studenttable
```

a) For an individual student their standing is ordinal; i.e. I moved up one grade. 

b) For the university itself the number of students within a given student standing is categorical. 

c) Because this is categorical data, I would prefer to use a bar chart. 

```{r}
#building up to bar chart by making a relative frequency distribution

n <- sum(studenttable)

#divide by sum and make new table w/ relative frequencies 

studentfreq <- studenttable/n
studentfreq
```

```{r}
#barplot
barplot(studenttable, col = c('red', 'green', 'blue', 'orange'),
        xlab = 'Student Standing',
        ylab = 'Count',
        ylim = c(0, 2000), main = 'Bar Graph of Student Standing Data')
```
## continuous data from contacts 

11) a. Collected some height data from a sample of 8 friends. 

```{r}
#entered heights in total inches 
heights <- c(70, 65, 66, 66, 68, 65, 62, 72)
```

I would say the heights are normally distributed with a left skew. Below are the min, max, median, and mean of this heights data. 

```{r}
summary(heights)
```

```{r}
IQR(heights) #IQR is 3.5 inches
sd(heights) #3.150964 is standard deviation 
```

b) The theoretical distribution is normal. 

c) Thinking of last semester, the normal distribution is conjugate to the normal distriubtion. 


## Chapter 4 Homework 

1) A conditional distribution is the distribution of observations of one variable given the values of the observationso f another variable. 

2) a. Based on visual inspection, as income increases, the values on the depression scale decreases.

b. The graph does not indicate that lower income causes depression since this negative relationship between increasing income and a depression scale does not explain the why or how of this relationship. A theoretical framework explaining how income relates to depression and additional data and data analysis is needed to say if there is a causal relationship rather than just an association.

3) I'll first create a table with the given data.

```{r}
#using as.table() and other functions to create a contingency table of counts. Source: anvas.evergreen.edu/courses/3164/modules/items/311641  

tab <- as.table(rbind(c(508, 349, 425, 288), c(371, 337, 384, 338)))
dimnames(tab) <- list(finaid = c("Yes", "No"),
                      class = c("First Year", "Sophomore", "Junior", "Senior"))
tab
```

```{r}
total <- rowSums(tab) #finding row sums
tab <- cbind(tab, total) #adding row sums to table
total <- colSums(tab) #reusing "total" for finding column sums 
fintab <- rbind(tab, total) #second table with sums for cross tabulation 
fintab #viewing table with totals 
```

```{r}
prop.table(fintab) #using prop.table 
```



