---
title: "Soc. 723 HW due 1.16.22"
author: "Brenda Onyango"
date: "1/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 3 Homework

1) What is a variable? A variable is a characteristic that takes on different values across a unit of analysis. In the chapter, monthly incomes of 433 South Africans was given as a variable, but it would be more precise to say that monthly income is the variable measured at the unit of the individual and that there are 433 observations for this variable. 

2) For each of the following variables describe if it is continuous, count, ordinal, categorical or qualitative. 

a) Age - continuous*
b) gender - categorical
c) number of times of presidential tweets - count 
d) income - continuous 
e) number of Insta posts about stats - count
g) college students' attend - categorical
h) therapist's written assessment - qualitative 
i) soccer divisions - ordinal 

*Technically continuous but I know practically for data analyses age is often treated as discrete. 

3) distribution

### Frequency Table and Salaries

Before answering 4, I created an object with given data to do some visualization and calculate relative frequencies. 

```{r}
#create table with salaries and frequency

salaryset <- c(rep('85K', 5), rep('90K', 4), rep('100K', 1), rep('120K', 2), 
rep('125K', 3), rep('130K', 2))

salarytable <- table(salaryset)

```

Let's look at the table.
```{r}
salarytable #notice that R arranged in different order than what was entered in set 
```

```{r}
#constructing relative frequency distribution of the data in the table.

s <- sum(salarytable)

#divide by sum to get relative frequencies
relfreq <- salarytable/s 

#view freq
relfreq
```

```{r}
dotchart(sort(relfreq),
         xlab = 'Relative frequency of salaries',
         main = 'Dot Chart of Econ Professor Salaries',
         xlim = c(0, 0.4), pch = 20, color = 'black')
```

a) Calculate average salary.
```{r}
salaries <- c(rep(85000, 5), rep(90000, 4), rep(100000, 1), rep(120000, 2), 
rep(125000, 3), rep(130000, 2))

head(salaries, 9) #confirming data entered correctly. Note that salaries is numerical and salaryset is categorical.
```

```{r}
mean(salaries)
```

Average salary is $103,529.40.

b) Find median.
```{r}
median(salaries)
```

The median is $90,000. 

c) Find min and max. 
```{r}
min(salaries)
max(salaries)
```
Min is 85,000 & max is $130,000. Instead of using min() and max(), I could have looked at the table. 

d) Find IQR. 
```{r}
IQR(salaries) #IQR is 40,000
summary(salaries) #quartiles 
```

5) c. Greek letters

6) a. Log transforming the variable 

7) Order from *low* to *high* variability: D, A, C, B  

8) 

a) The distribution ranges from about 73 about 89 with a mode of about 79. 

b) There is a slight right skew. 

c) Because of the right skew (very few people scoring very well), the median may be a better measure for the center of distribution.

d) I would use standard deviation to describe variability. 

9) b. The mean< median. 

### Frequency Table and Students 

```{r}
#create table with grade and frequency

students <- c(rep('First Year', 1000), rep('Sophomore', 1200), rep('Junior', 900), rep('Senior', 1500))

studenttable <- table(students)
studenttable
```

a) For an individual student their standing is ordinal; i.e. I moved up one grade. 

b) For the university itself the number of students within a given student standing is categorical. 

c) Because this is categorical data, I would prefer to use a bar chart. 

```{r}
#building up to bar chart by making a relative frequency distribution

n <- sum(studenttable)

#divide by sum and make new table w/ relative frequencies 

studentfreq <- studenttable/n
studentfreq
```

```{r}
#barplot
barplot(studenttable, col = c('red', 'green', 'blue', 'orange'),
        xlab = 'Student Standing',
        ylab = 'Count',
        ylim = c(0, 2000), main = 'Bar Graph of Student Standing Data')
```

## continuous data from contacts 

11) a. Collected some height data from a sample of 8 friends. 

```{r}
#entered heights in total inches 
heights <- c(70, 65, 66, 66, 68, 65, 62, 72)
```

I would say the heights are normally distributed with a left skew. Below are the min, max, median, and mean of this heights data. 

```{r}
summary(heights)
```

```{r}
IQR(heights) #IQR is 3.5 inches
sd(heights) #3.150964 is standard deviation 
```

b) The theoretical distribution is normal. 

c) Thinking of last semester, the normal distribution is conjugate to the normal distriubtion. 


## Chapter 4 Homework 

1) A conditional distribution is the distribution of observations of one variable given the values of the observationso f another variable. 

2) a. Based on visual inspection, as income increases, the values on the depression scale decreases.

b. The graph does not indicate that lower income causes depression since this negative relationship between increasing income and a depression scale does not explain the why or how of this relationship. A theoretical framework explaining how income relates to depression and additional data and data analysis is needed to say if there is a causal relationship rather than just an association.

3) I'll first create a table with the given data.

```{r}
#using as.table() and other functions to create a contingency table of counts. Source: anvas.evergreen.edu/courses/3164/modules/items/311641  

tab <- as.table(rbind(c(508, 349, 425, 288), c(371, 337, 384, 338)))
dimnames(tab) <- list(finaid = c("Yes", "No"),
                      class = c("First Year", "Sophomore", "Junior", "Senior"))
tab
```

```{r}
total <- rowSums(tab) #finding row sums
tab <- cbind(tab, total) #adding row sums to table
total <- colSums(tab) #reusing "total" for finding column sums 
fintab <- rbind(tab, total) #second table with sums for cross tabulation 
fintab #viewing table with totals 
```

```{r}
#3A
#finding conditional probability of p(yes|senior)
fintab[1,4]/fintab[3,4] #numerator is 1st row fourth column and denominator is 3rd row fourth column 
```

a) P(finaid Yes|senior) = 0.46

```{r}
#3B
#finding conditional probability of p(senior|yes)
fintab[1,4]/fintab[1,5]
```

b) p(senior|finaid Yes)

```{r}
#3C
#finding conditional probability of p(yes|first year)
fintab[1,1]/fintab[3,1] #first years with finaid/total number of first years
```

c) p(finaid yes|first year) = 0.5779

4) Two advantages of line-fitting methods are that 1) we can get the conditional mean of Y for any value of X even when we don't have data and 2) that we can describe the relationship between X and Y in a more precise w way than methods using local means. One disadvantage of line-fitting is needing to select the correct line to use (e.g. is the line of best fit linear or parabolic). 

5) 
```{r}
#plotting the given line source: http://www.countbio.com/web_pages/left_object/R_for_biology/R_fundamentals/curve_to_function_R.html
curve(4 - 3.5*x, from=-10, to=10, n = 25, xlab="x value", ylab = "y value", 
      col ="blue", lwd=3, main = "Plot of Y= 4 - 3.5x")
```
 Reminder to self that conditional mean is asking given a certain value of X, what do I expect the mean of Y to be? 
 
 a) Y = 4 - 3.5(6) = -17
 b) Y = 4 - 3.5(-2) = 11
 
 6. c. covariance 
 
 7. Covariance shows to what extent two variables move together or apart, and has a theoretical range of values between negative and positive infinity. A positive covariance means that an increase in one variable would mean that the other variable increases. A covariance of 0 means that the two variables don't vary with each other. Covariance only tells us the direction, if any, of the relationship between variables.
 
 Pearson's correlation coefficient can only range between -1 and 1 and tells us the direction *and* strength of the relationship between two variables. Another difference between covariance and correlation is that correlation is standardized (which is why it can only take values between -1 to 1). 
 
 8.
 a) A is fitted to a line.
 
 b) B is fitted to what appears to be a LOESS curve or a parabola. 
 
 c) B is a better fit for the data and I can tell because the residuals for B are smaller than for A. 
 
 d) For exam scores between -2 and -1, the residuals are mostly positive (above the curve). For exam scores between -1 and 1 residuals are mostly negative (below the curve). The residuals are mostly below the curve as most data points are below the line. 
 
 9) 
 
 a) 
 
```{r}
#observed - prediction = residual 
observed <- c(5.07, 4.05, 5.33, 4.02, 3.59) #creating vectors to find residuals 
prediction <- c(4.72, 4.28, 5.38, 4.22, 3.45)
observed - prediction
```


```{r}
#loading library to use texttotable
library(DescTools)
```

 
```{r}

#recreating table using #txttotable
txt <- "GetAlong, Satisfaction, Prediction, Residual
    4.7, 5.07, 4.72,  0.35
    4.21, 4.05, 4.28, -0.23
    5.42, 5.33, 5.38, -0.05
    4.14, 4.02, 4.22, -0.20
    3.3, 3.59, 3.45,  0.14"
TextToTable(txt, sep=",", dimnames = c(" ", "Coworker Congeniality and Satisfaction Report"))

#QUESTION to Nico: how do I get rid of the V1, V2... and A,B, C, ...?
```

 b) When fitting a line with ordinary least squares, the line that gives the lowest sum of squared residuals is the best fit. 
 
 10) 
```{r}
library("tidyverse")
library("DiagrammeR")
```
 
```{r}
tree <-
    create_graph() |>   # creating tree graph to represent causal diagram
    add_n_nodes( #adding nodes on tree
      n = 3, 
      type = "path",
      label = c("familial poverty", "first generation status", "graduation rate"), # Labelling nodes
      node_aes = node_aes(
        shape = "rectangle",
        height = 1.5,
        width = 1.5,
        x = c(0, 3, 3), # specifying height of nodes
        y = c(0, 2, -2), 
        fill = c("black", "blue", "orange"))) |>  #coloring nodes 
    add_edge(
      from = 1,
      to = 2,
      edge_aes = edge_aes(
        label = "+" 
      )) |>  
  add_edge(
      from = 1,
      to = 3,
      edge_aes = edge_aes(
        label = "-"
      )) 
  
       
render_graph(tree)
```
 
 a) Poverty measured by family income could explain both first generation status and graduation rate. Poverty could lead to a decreased graduate rate if students have barriers to registering and continuing coursework due to needing to work or high bursar balances. Familial poverty in areas where college costs are high can explain a high proportion of first generation students. This diagram could apply to high school first generation status and high school graduation rates, too. 
 
 b) 
 1) Get the mean of graduation rate conditional on poverty. (Given poverty value what is graduation rate?)
 
 2) Subtract the conditional mean to get the residual of graduation rate.
 
 3) Get the mean of first generation status conditional on poverty. (Given poverty value what is first generation status?)
 
 4) Subtract the conditional mean to get the residual of first generation status. 
 
 5) Describe the relationship betweeen the residual of graduation rate and the residual of first generation status. 
 
 c) I would interpret the result from the above steps as the relationship between first generation status and graduation rate that is *not* explained by familial poverty. It is the mean of graduation rate conditional on first generation status when poverty doesn't vary. 